Likelihoods for a general class of ARGs under the SMC

### 2025-02-21 paper.tex
edits

In Fig~\ref{fig:lik-surface} we begin with a realisation of the SMC 
containing complete information about every simulated event
(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). 
We then plot the likelihood computed by Eq.~\eqref{eq:full-lik} 
as we vary the recombination rate parameter $r$, as a ratio
of the likelihood at the true value of $r$. 
Reassuringly, we can
see that the minimum of the likelihood-ratio curve when using the 
full simulated ARG is close to the true value of the parameter.
(The likelihood ratio curve obtained when computing the likelihood
under the full coalescent with recombination model using the KYF
formula is identical; not shown.)
We next compute the likelihood ratio for the ARG obtained when we 
simplify out the recombination nodes from the full ARG
(RE nodes removed; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime 
uses distinct nodes for the left and right parent of each recombination).
The curve we obtain is indistinguishable from that of the full
ARG, demonstrating that the locally unary portions of coalescent nodes 
contain all the information that we need about recombination events
for this application.
When we fully simplify the simulated ARG
to remove all locally unary nodes from trees 
(Fully simplified; 3,332 nodes; 10,674 edges, 2,933 trees) 
we can see there is a significant loss of information,
and the minimum of the 
likelihood ratio surface is biased away from its true value.
(See~\cite{wong_general_2023} for more discussion on varying 
degrees of ARG simplification.)

Fig~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably
robust to the presences of polytomies in this example. For each of the three
ARGs discussed in the previous paragraph, we also evaluated the likelihood
for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and 
edges adjusted accordingly.
This introduces polytomies at varying degrees in the different ARGs because 
deleting a node in a tree (and connecting its children directly to its
parent) will only create a polytomy if the parent 
is not unary. Thus, while deleting 25\% of the internal nodes 
in the fully simplified ARG created an average of 33 polytomies
over the 2,933 trees along the sequence (each tree having 
200 leaf nodes), it resulted in an average of 21 polytomies
per tree in the ARG where recombination nodes have been removed,
and only 10 in the full ARG (which has an average of 1026 unary 
nodes per tree).
%     name                  leaves  unary       binary  poly
% 0   full_arg_smc          200.0   1094.417661 199.000000  0.000000
% 1   unary_coal            200.0   456.396522  199.000000  0.000000
% 2   fully_simplified      200.0   0.000000    199.000000  0.000000
% 3   tsinfer_tsdate        200.0   192.447590  119.754399  32.270849
% 4   tsinfer_fs_tsdate     200.0   0.000000    119.563898  32.322684
% 5   py_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
% 6   py_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
% 7   py_fully_simied_0.05  200.0   0.000000    185.691468  6.393174
% 8   py_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
% 9   py_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
% 10  py_fully_simied_0.25  200.0   0.000000    117.096174  33.241986
Given this disparity in the number of polytomies introduced, the 
effects on the likelihood across the three examples are not entirely
comparable. For the ARGs containing unary nodes, 
removing 25\% of the internal 
nodes has negligible effect, and even on the fully simplified 
ARG has a very minor effect on location of the minimum.

Fig~\ref{fig:lik-surface} also shows results for ARGs inferred from the 
simulated data using tsinfer and tsdate~\citep{wohns_unified_2022},
using default parameters for tsinfer, and the true values of mutation
rate and $N_e$ for tsdate.
In the first, we use the default output of tsinfer 
(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations).
This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
with three or more children per tree. Tsinfer does not directly estimate
recombination events; the breakpoints between trees are the consequence of
switches between parental haplotypes in the 
\cite{li2003modeling} copying process. Information about 
recombination events in the returned ARG is therefore quite imprecise,
and likely to contain incorrect and contradictory details if one were to 
attempt to reconstruct the exact subtree-prune-and-regraft 
moves~\citep[e.g.][]{rasmussen2023espalier}. Nonetheless, in
this toy example at least, our likelihood function performs well
on the output of tsinfer, with the minimum close to the true
parameter value. We also show the likelihood curve obtained when we 
fully simplify the output of tsinfer 
(1,865 nodes; 7,511 edges; 1,252 trees; 3,039 mutations)
before dating. We can see that, as with the true data,
removing all the unary nodes results in a significant bias
away from the true value of the recombination rate parameter.

@@

\section*{Discussion}
The key contribution of this paper is to rederive a classical
ARG likelihood function in a way that supports the sort of partial 
and imprecise information
provided by some modern ARG inference tools,
using a novel backwards-time derivation of the SMC.
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, because we
do not necessarily observe or estimate these events,
we integrate over the possible timings implied by the information
in the input ARG.
Our method can compute a likelihood under the SMC for a very general class of 
ARG~\citep{wong_general_2023}, 
a significant increase in flexibility over the 
strict input requirements for the classical Kuhner-Yamato-Felsenstein 
formulation. 
Locally unary nodes play a key role, and although not
currently a common focus of ARG research~\citep{wong_general_2023}, 
they are estimated by tsinfer and can be imputed to high a
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
We have shown via some illustrative examples that the formulation is 
robust to incomplete and imprecise information, tolerating the loss 
of a significant fraction of the internal nodes in the ARG to minimal effect.
We have also shown that the likelihood function performs well 
(at least in our toy example) on ARGs inferred by tsinfer+tsdate,
successfully capturing information about latent recombination events
from imprecise and noisy data.
% I think we have to grasp this nettle or we're inviting the reviewers to 
% give us a bunch of "what if you tried this" bits of extra homework.
Theoretical explanations of this apparent robustness,
and fully characterising the properties of the likelihood on the output 
of tsinfer and other methods is an important avenue for future work.
Similarly, while there are many potential applications to concrete 
inferential problems, the details are nontrivial and must be
consigned to future work.

>>>>

In Fig~\ref{fig:lik-surface} we begin with a realisation of the SMC 
containing complete information about every simulated event
(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). 
We then plot the likelihood computed by Eq.~\eqref{eq:full-lik} 
as we vary the recombination rate parameter $r$, as a ratio
of the likelihood at the true value of $r$. 
Reassuringly, we can
see that the minimum of the likelihood-ratio curve when using the 
full simulated ARG is close to the true value of the parameter.
(The likelihood ratio curve obtained when computing the likelihood
under the full coalescent with recombination model using the KYF
formula is identical; not shown.)
We next compute the likelihood ratio for the ARG obtained when we 
simplify out the recombination nodes from the full ARG
(``RE nodes removed''; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime 
uses distinct nodes for the left and right parent of each recombination).
The curve we obtain is indistinguishable from that of the full
ARG, demonstrating that the locally unary portions of coalescent nodes 
contain all the information that we need about recombination events
for this application.
When we fully simplify the simulated ARG
to remove all locally unary portions of nodes from trees 
(``Fully simplified''; 3,332 nodes; 10,674 edges, 2,933 trees) 
we can see there is a significant loss of information,
and the minimum of the 
likelihood ratio surface is biased away from its true value,
as expected, because removing the unary portions of nodes has removed some of the area
of the ARG in which recombination did not occur.
(See~\cite{wong_general_2023} for more discussion on varying 
degrees of ARG simplification.)

Fig~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably
robust to the presences of polytomies in this example. For each of the three
ARGs discussed in the previous paragraph, we also evaluated the likelihood
for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and 
edges adjusted accordingly.
This introduces polytomies at varying degrees in the different ARGs because 
deleting a node in a tree (and connecting its children directly to its
parent) will only create a polytomy if the parent 
is not unary. Thus, while deleting 25\% of the internal nodes 
in the fully simplified ARG created an average of 33 polytomies
over the 2,933 trees along the sequence (each tree having 
200 leaf nodes), it resulted in an average of 21 polytomies
per tree in the ARG where recombination nodes have been removed,
and only 10 in the full ARG (which has an average of 1026 unary 
nodes per tree).
%     name                  leaves  unary       binary  poly
% 0   full_arg_smc          200.0   1094.417661 199.000000  0.000000
% 1   unary_coal            200.0   456.396522  199.000000  0.000000
% 2   fully_simplified      200.0   0.000000    199.000000  0.000000
% 3   tsinfer_tsdate        200.0   192.447590  119.754399  32.270849
% 4   tsinfer_fs_tsdate     200.0   0.000000    119.563898  32.322684
% 5   py_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
% 6   py_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
% 7   py_fully_simied_0.05  200.0   0.000000    185.691468  6.393174
% 8   py_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
% 9   py_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
% 10  py_fully_simied_0.25  200.0   0.000000    117.096174  33.241986
Given this disparity in the number of polytomies introduced, the 
effects on the likelihood across the three examples are not entirely comparable.
For the ARGs containing unary nodes, 
removing 25\% of the internal 
nodes has negligible effect on the likelihood curve in Fig~\ref{fig:lik-surface},
and even on the fully simplified 
ARG has only a very minor effect on location of the minimum.

Fig~\ref{fig:lik-surface} also shows results for ARGs inferred from the 
simulated data using tsinfer and tsdate~\citep{wohns_unified_2022},
using default parameters for tsinfer, and the true values of mutation
rate and $N_e$ for tsdate.
In the first, we use the default output of tsinfer 
(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations).
This ARG has 192 unary nodes, 120 binary nodes, and 32 nodes
with three or more children per tree. Tsinfer does not directly estimate
recombination events; the breakpoints between trees are the consequence of
switches between parental haplotypes in the 
\cite{li2003modeling} copying process. Information about 
recombination events in the returned ARG is therefore quite imprecise,
and likely to contain incorrect and contradictory details if one were to 
attempt to reconstruct the exact subtree-prune-and-regraft 
moves~\citep[e.g.,][]{rasmussen2023espalier}. Nonetheless, in
this toy example at least, our likelihood function performs well
on the output of tsinfer, with the minimum close to the true
parameter value. We also show the likelihood curve obtained when we 
fully simplify the output of tsinfer 
(1,865 nodes; 7,511 edges; 1,252 trees; 3,039 mutations)
before dating. We can see that, as with the true data,
removing all the unary nodes results in a significant bias
away from the true value of the recombination rate parameter.

@@

\section*{Discussion}
The key contribution of this paper is to rederive a classical
ARG likelihood function in a way that supports the sort of partial 
and imprecise information
provided by some modern ARG inference tools,
using a novel backwards-time derivation of the SMC.
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, because we
do not necessarily observe or estimate these events,
we integrate over the possible timings implied by the information
in the input ARG.
Our method can compute a likelihood under the SMC for a very general class of 
ARG~\citep{wong_general_2023}, 
a significant increase in flexibility over the 
strict input requirements for the classical Kuhner-Yamato-Felsenstein 
formulation. 
Locally unary nodes play a key role, and although not
currently a common focus of ARG research~\citep{wong_general_2023}, 
they are estimated by tsinfer and may be imputed to a high
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
We have shown via some illustrative examples that the formulation is 
robust to incomplete and imprecise information, tolerating the loss 
of a significant fraction of the internal nodes in the ARG to minimal effect.
We have also shown that the likelihood function performs well 
(at least in our toy example) on ARGs inferred by tsinfer+tsdate,
successfully capturing information about latent recombination events
from imprecise and noisy data.
However, there are many more aspects of the likelihood function
that could be examined, beyond coarse parameters like recombination rate
or effective populations size.
% I think we have to grasp this nettle or we're inviting the reviewers to 
% give us a bunch of "what if you tried this" bits of extra homework.
Theoretical explanations of this apparent robustness,
and fully characterising the properties of the likelihood on the output 
of tsinfer and other methods is an important avenue for future work.
Similarly, while there are many potential applications to concrete 
inferential problems, the details are nontrivial and must be
consigned to future work.
